Value-Suppressing Uncertainty Maps

### 2017-03-30 paper/template.tex
Typo
Add example figures

@@

In this paper, we contribute \textbf{Value-Suppressing Uncertainty Maps} (VSUMs) for integrating data and uncertainty information in thematic maps. VSUMs intentionally alias together data values with high uncertainty, affording greater discriminability as uncertainty decreases. Rather than a traditional two-dimensional bivariate map, with differing outputs for each combination of value and uncertainty, VSUMs can be thought of as arcs: as uncertainty increases, values are mapped to smaller and smaller sets of outputs, culminating in a singularity where all inputs are mapped to an identical, highly uncertain mark regardless of their data value. Figure~\ref{fig:example} shows examples of both a VSUM and a more traditional bivariate map.

@@

Given these considerations, and the perceptual integrality of color channels, the construction of \emph{bivariate} color maps is difficult. Viewers need to be able to independently assess the variables of interest, but the resulting bivariate map should also be expressive and easy to interpret. Bivariate maps with these constraints are often limited to relatively few categories (say, a 3x3 matrix as in Figure \ref{fig:example}) \cite{robertson1986generation,trumbo1981theory}. In general, the quality of bivariate color maps is a multivariate measure involving consideration of not just the component color channels, but also the interpolation scheme and the color of the surround~\cite{bernard2015survey}. However, even simple bivariate maps can be difficult for a general audience to interpret \cite{wainer1980empirical}.

@@

\begin{enumerate}
	\item \textbf{Select a visual variable to encode value}. In most cases in this paper, this variable is position along some established color scale.
	\item \textbf{Select a visual variable to encode uncertainty}. We prefer variables that have a somewhat high semiotic connection with uncertainty (such as lightness and size), but still have a reasonable number of perceptually discriminable layers (as opposed to other, more intuitive channels such as blur or sketchiness).
	\item \textbf{Select a discriminability threshold}. Choose some minimum distance, in a perceptual space, between marks. For many visual variables (such as length or hue), the perceptual discriminability between marks has been assessed in prior literature, and so the only step is to choose a value in terms of the just noticeable distance (JND). For less common visual variables, a perceptual model may be built empirically (as in Demiralp et al.~\cite{demiralp2014learning}).
	\item \textbf{Choose how quickly categories \emph{degrade}}. That is, if there are $n$ bins at some uncertainty level $m$, how many bins should there be when uncertainty increases to level $m+1$? In Figure \ref{fig:example}, this function is $\left \lfloor {\frac{n}{2}}\right \rfloor$, meaning that the number of value bins decreases by half for each level of uncertainty bins. If there is too much degradation, then there are fewer levels of uncertainty. If there is no degradation, the resulting map is just a traditional bivariate map.
	\item \textbf{Choose the largest number of initial categories $n$ such that the discriminability threshold is met.} For each pair of marks in the map, perform a pairwise comparison of discriminability. If no marks are perceptually ambiguous by the standards of your threshold, the map is complete. Otherwise, decrement $n$.
\end{enumerate}

@@

We use values from these works as benchmarks for selecting thresholds of discriminability. The ``just noticeable difference'' (JND) of these channels provides a useful minimum; in practice, we would wish to create maps with marks that are far more reliably distinguishable than the 50\% or 75\% discriminability rate of a JND. Figure~\ref{fig:performance} shows how these benchmarks can be used to select reasonable thresholds. Other factors, such as nameability and aesthetic preference, can be added in as additional weights impacting the desirability of certain color palettes, as in Gramazio et al.~\cite{gramazio2017colorgorical}.

@@

TODO.

@@

We included both a \textbf{gain} framing (given predictions of where your opponent's ships are, where should you launch your missiles?), and a \textbf{loss} framing (given predictions of where your opponent's missiles are, where should you place your ships?) in order to account for systematic differences in human judgments under uncertainty. Tversky \& Kahneman~\cite{tversky1985framing} illustrate that framings in terms of gains or losses produce reliably different outcomes. In general, the function of the value of gains is concave, but is convex for losses~\cite{kahneman1979prospect}. That is, avoiding a large loass is often more valuable than striving for a similarly large gain.

>>>>

\newcommand{\figref}[1]{\hyperref[#1]{Figure~\ref*{#1}}}

\newcommand{\ie}{{i.e.}\xspace}
\newcommand{\eg}{{e.g.,}\xspace}
\newcommand{\cf}{{c.f.}\xspace}
\newcommand{\ea}{{et~al.}\xspace}
\newcommand{\aka}{{a.k.a.}\xspace}
\newcommand{\etc}{{etc.}\xspace}

@@

In this paper, we contribute \textbf{Value-Suppressing Uncertainty Maps} (VSUMs) for integrating data and uncertainty information in thematic maps. VSUMs intentionally alias together data values with high uncertainty, affording greater discriminability as uncertainty decreases. Rather than a traditional two-dimensional bivariate map, with differing outputs for each combination of value and uncertainty, VSUMs can be thought of as arcs: as uncertainty increases, values are mapped to smaller and smaller sets of outputs, culminating in a singularity where all inputs are mapped to an identical, highly uncertain mark regardless of their data value. \figref{fig:example} shows examples of both a VSUM and a more traditional bivariate map.

@@

Given these considerations, and the perceptual integrality of color channels, the construction of \emph{bivariate} color maps is difficult. Viewers need to be able to independently assess the variables of interest, but the resulting bivariate map should also be expressive and easy to interpret. Bivariate maps with these constraints are often limited to relatively few categories (say, a 3x3 matrix as in \figref{fig:example}) \cite{robertson1986generation,trumbo1981theory}. In general, the quality of bivariate color maps is a multivariate measure involving consideration of not just the component color channels, but also the interpolation scheme and the color of the surround~\cite{bernard2015survey}. However, even simple bivariate maps can be difficult for a general audience to interpret \cite{wainer1980empirical}.

@@

\begin{enumerate}
	\item \textbf{Select a visual variable to encode value}. In most cases in this paper, this variable is position along some established color scale.
	\item \textbf{Select a visual variable to encode uncertainty}. We prefer variables that have a somewhat high semiotic connection with uncertainty (such as lightness and size), but still have a reasonable number of perceptually discriminable layers (as opposed to other, more intuitive channels such as blur or sketchiness).
	\item \textbf{Select a discriminability threshold}. Choose some minimum distance, in a perceptual space, between marks. For many visual variables (such as length or hue), the perceptual discriminability between marks has been assessed in prior literature, and so the only step is to choose a value in terms of the just noticeable distance (JND). For less common visual variables, a perceptual model may be built empirically (as in Demiralp et al.~\cite{demiralp2014learning}).
	\item \textbf{Choose how quickly categories \emph{degrade}}. That is, if there are $n$ bins at some uncertainty level $m$, how many bins should there be when uncertainty increases to level $m+1$? In \figref{fig:example}, this function is $\left \lfloor {\frac{n}{2}}\right \rfloor$, meaning that the number of value bins decreases by half for each level of uncertainty bins. If there is too much degradation, then there are fewer levels of uncertainty. If there is no degradation, the resulting map is just a traditional bivariate map.
	\item \textbf{Choose the largest number of initial categories $n$ such that the discriminability threshold is met.} For each pair of marks in the map, perform a pairwise comparison of discriminability. If no marks are perceptually ambiguous by the standards of your threshold, the map is complete. Otherwise, decrement $n$.
\end{enumerate}

@@

We use values from these works as benchmarks for selecting thresholds of discriminability. The ``just noticeable difference'' (JND) of these channels provides a useful minimum; in practice, we would wish to create maps with marks that are far more reliably distinguishable than the 50\% or 75\% discriminability rate of a JND. \figref{fig:performance} shows how these benchmarks can be used to select reasonable thresholds. Other factors, such as nameability and aesthetic preference, can be added in as additional weights impacting the desirability of certain color palettes, as in Gramazio et al.~\cite{gramazio2017colorgorical}.

@@

\airlineFig
\viralFig

\figref{fig:airlineVsum}
\figref{fig:airline2d}

@@

We included both a \textbf{gain} framing (given predictions of where your opponent's ships are, where should you launch your missiles?), and a \textbf{loss} framing (given predictions of where your opponent's missiles are, where should you place your ships?) in order to account for systematic differences in human judgments under uncertainty. Tversky \& Kahneman~\cite{tversky1985framing} illustrate that framings in terms of gains or losses produce reliably different outcomes. In general, the function of the value of gains is concave, but is convex for losses~\cite{kahneman1979prospect}. That is, avoiding a large loss is often more valuable than striving for a similarly large gain.
